---
title: "Covid Rough Draft"
author: "Data Humanist"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes 
    toc_float: yes
  toc: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	fig.fullwidth = TRUE,
	message = FALSE,
	warning = FALSE
)
```


```{r, libraries_data}
library(tidyverse)
library(here)
library(visdat)
library(reactable)
library(glue)


here()

load(here::here("tidy_data", "project_data.RData"))

# Custom breaks for the plots to come -------------------------------------

breaks_sc <- c(seq(0, 0.2, by = 0.05), .25, .50, .75)
breaks_sc1 <- c(seq(0, 20, by = 5), 25, 50, 75)
breaks_per <- seq(0.00, 0.14, by = 0.02)
breaks_per2 <- seq(0.00, 0.22, by = 0.02)

cust_cap <- "Replace with appropriate caption"

##
# coutesy of https://www.r-bloggers.com/2019/12/vignette-downloadable-tables-in-rmarkdown-with-the-dt-package/

create_dt <- function(x){
  DT::datatable(x,
                caption = cust_cap,
                extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                buttons = c('copy', 'csv', 
                            'excel', 'pdf', 'print'),
                lengthMenu = list(c(10,25,50,-1),  c(10,25,50,"All"))
                
                ) )
}


```

## Brief

This RMD generates the images and data tables. Primary data source: USA CDC, [Provisional Death Counts for COVID-19](https://www.cdc.gov/nchs/nvss/vsrr/covid_weekly/index.htm).  Retrieved 19 January 2022. Deaths counts might lag slightly. Please read the CDC Disclaimer. 

The clean dashboard version, with better graphics @ [Covid in the USA: 2021 vs. 2020](https://rpubs.com/Thom_JH/Covid_Relative_Risk)

Polemic commentary on these results at Substack: [American Exile](https://americanexile.substack.com/).

If you make use of this data or worksheet, please consider: http://buymeacoffee.com/datahumanist

Thank you!


Thank you.



## Big Picture



### Plots All
```{r}

all_chart <- All_Sexes_long %>%
  mutate(Percentages = Ratio_Vals * 100) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  filter(Deaths_Type %in% c("With Covid","No Covid"))  %>%
  ggplot( aes(x = Age_Range,  y = Percentages ,  fill = Deaths_Type) ) +
  geom_bar(position="stack", stat="identity") +
  facet_grid(Year ~ .) +
  coord_flip() +
  theme(legend.position = c(0.5, 0.5) , 
        legend.background = element_rect(linetype = 2))+
  scale_y_continuous( breaks =  breaks_sc1) +
  scale_fill_manual(values = c("#121121", "#FF4500") ) +
  labs(title = "All Sexes: Covid vs. Non-Covid Deaths",
       fill = "Death Types", x = "Age Ranges", y= "Percentages",
       subtitle = "US CDC data for years 2021 & 2020",
       caption = "Data Humanist, CC BY-SA 4.0") +
  geom_hline(yintercept = breaks_sc1 , color = "white", lty = 3) +
  geom_text( aes(label = (({Percentages}) %>% round(1)) ), size = 2.2, 
             vjust = "inward", hjust = "inward", color= "white") 



all_chart

```



```{r}
rr_all <- Comp_2021_vs_2020_ALL %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  ggplot(aes(x = Age_Range,  y = Relative_Risk,  fill =  Relative_Risk)) +
  geom_col() +
  scale_fill_viridis_c(option = "D") +
  guides(fill = "none")  +
  labs(title = "Relative Risk: Dying with Covid in 2021 vs. 2020",
       subtitle = "US CDC data. Source stratified by Age Range.",
       caption = "Data Humanist, CC BY-SA 4.0",
       x = "Age Range", y = "Relative Risk") +
  geom_hline(yintercept = 1, color = "white", lty = 3) +
  geom_text( aes(label = paste0(round({Relative_Risk}, 2), "x" ) ), size = 3, 
             vjust = 1.5, color= "red") 


rr_all 
```



### Data Sets ALL


### Stats_All

#### reactable

```{r}

All_Sexes %>% 
  group_by( Age_Range, Year)  %>%
  summarise("Deaths with C19" = Deaths_with_Covid,
            "Deaths All Causes" = Death_All_Causes,
            "C19 to All Deaths (%)" = (C19_Ratio * 100) %>% round(2) ) %>%
  reactable(., highlight = TRUE, 
            striped = TRUE,
            pageSizeOptions = c(10, 25, 50, 100),
            theme = reactableTheme(
              stripedColor = "#EDEDED",
              highlightColor = "#FFE4E1") ) 


```


#### knitr

```{r}
All_Sexes %>% 
  group_by( Age_Range, Year)  %>%
  summarise("Deaths with C19" = Deaths_with_Covid,
            "Deaths All Causes" = Death_All_Causes,
            "C19 to All Deaths (%)" = (C19_Ratio * 100) %>% round(2) )

```


#### DT

```{r}
cust_cap <- "CDC Data: Provisional COVID-19 Death Counts, 2020-2021: All Sexes"
All_Sexes %>% 
  group_by( Age_Range, Year)  %>%
  summarise("Deaths with C19" = Deaths_with_Covid,
            "Deaths All Causes" = Death_All_Causes,
            "C19 to All Deaths (%)" = (C19_Ratio * 100) %>% 
              round(2) ) %>% create_dt()
```


Note: going with DT.












### Count Plot


```{r}
all_count <- All_Sexes_long %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  filter(Deaths_Type %in% c("With Covid","No Covid"))  %>%
  ggplot( aes(x = Age_Range,   y= Deaths_Count,  fill = Deaths_Type) ) +
  geom_bar(position="stack", stat="identity") +
  labs(title = "All Sexes: Covid vs. Non-Covid Deaths",
       fill = "Death Types", x = "Age Ranges", 
       subtitle = "US CDC data for years 2021 & 2020",
       caption = "Data Humanist, CC BY-SA 4.0",
       y = "Death Count") +
  facet_grid(Year ~ .) +
  coord_flip() +
  theme(legend.position = c(0.8, 0.15)) +
  scale_fill_manual(values = c("#121121", "#FF4500")) 

all_count 
```

Note: count plots widely available online, and this one -- like those -- obscures the changes in percentages and relative risk.

## By Sex


### Female

```{r}


female_chart <- My_data_long %>% 
  mutate(Percentages = Ratio_Vals * 100) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  filter(Deaths_Type %in% c("With Covid","No Covid"))  %>%
  filter(Sex == "Female") %>% 
  ggplot( aes(x = Age_Range,  
              y = Percentages,  
              fill = Deaths_Type) ) +
  geom_bar(position="stack", stat="identity") +
  facet_grid(Year ~ .) +
  scale_y_continuous(breaks = breaks_sc1 ) +
  coord_flip() +
  theme(legend.position = c(0.5, 0.5)) +
  scale_fill_manual(values = c("#696969", "#FF4500") ) +
  labs(title = "Female: Deaths w/ Covid vs. Non-Covid",
       subtitle = "US CDC data for years 2021 & 2020",
       caption = "Data Humanist, CC BY-SA 4.0",
       y = "Percentages", x = "Age Range", fill = "Death Types")  +
  geom_hline(yintercept = breaks_sc , color = "#FFFFF0", lty = 3) +
  geom_text(aes(label = (({Percentages} ) %>% round(1)) ), size = 2.2, 
             vjust = "inward", hjust = "inward", color= "#F8F8FF") 


female_chart 


```




```{r}
rr_female <- Comp_2021_vs_2020 %>%
  filter(Sex == "Female") %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  ggplot(aes(x = Age_Range,  y = Relative_Risk,  fill =  Relative_Risk)) +
  geom_col() +
  scale_fill_viridis_c(option = "D") +
  guides(fill = "none")  +
  labs(title = "Female Relative Risk: Dying with Covid in 2021 vs. 2020",
       subtitle = "US CDC data. Source stratified by Age Range.",
       caption = "Data Humanist, CC BY-SA 4.0",
       x = "Age Ranges", y = "Relative Risk") +
  geom_hline(yintercept = 1, color = "white", lty = 3) +
  geom_text( aes(label = paste0(round({Relative_Risk}, 2), "x" ) ), size = 3, 
             vjust = 1.5, color= "red") 

rr_female 
```





```{r}

cust_cap <- "CDC Data for Female Sex: Provisional Death Counts for COVID-19" 

Female_DT <-  My_data %>% 
  group_by(Age_Range, Year)  %>%
  filter(Sex == "Female") %>%
  summarise("Deaths with C19" = Deaths_with_Covid,
            "Deaths All Causes" = Death_All_Causes,
            "C19 to All Deaths (%)" = (C19_Ratio * 100) %>% round(2) )

create_dt(Female_DT) 
  



```


### Male

```{r}
breaks_sc1 <- c(seq(0, 20, by = 5), 25, 50, 75)
male_chart <- My_data_long %>% 
  mutate(Percentages = Ratio_Vals * 100) %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  filter(Deaths_Type %in% c("With Covid","No Covid"))  %>%
  filter(Sex == "Male") %>% 
  ggplot( aes(x = Age_Range,  
              y = Percentages ,  
              fill = Deaths_Type) ) +
  geom_bar(position="stack", stat="identity") +
  facet_grid(Year ~ .) +
  scale_y_continuous(breaks = breaks_sc1 ) +
  coord_flip() +
  theme(legend.position = c(0.5, 0.5)) +
  scale_fill_manual(values = c("#4682B4", "#FF4500")) +
  labs(title = "Male: Deaths w/ Covid vs. Non-Covid",
       subtitle = "US CDC data for years 2021 & 2020",
       caption = "Data Humanist, CC BY-SA 4.0",
       y = "Percentages", x = "Age Range", fill = "Death Types") +
  geom_hline(yintercept = breaks_sc1 , color = "white", lty = 3) +
  geom_text( aes(label = (({Percentages}) %>% round(1)) ), size = 2.2, 
             vjust = "inward", hjust = "inward", color= "#F8F8FF") 

male_chart

```



```{r}


rr_male <- Comp_2021_vs_2020 %>%
  filter(Sex == "Male") %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  ggplot(aes(x = Age_Range,  y = Relative_Risk,  fill =  Relative_Risk)) +
  geom_col() +
  scale_fill_viridis_c(option = "D") +
  guides(fill = "none")  +
  labs(title = "Male Relative Risk: Dying with Covid in 2021 vs. 2020",
       subtitle = "US CDC data. Source stratified by Age Range.",
       caption = "Data Humanist, CC BY-SA 4.0",
       x = "Age Ranges", y = "Relative Risk") +
  geom_hline(yintercept = 1, color = "white", lty = 3) +
  geom_text( aes(label = paste0(round({Relative_Risk}, 2), "x" ) ), 
             size = 3, 
             vjust = 1.5, color= "red") 

rr_male
```
```{r}
cust_cap <- "CDC Data for Male Sex: Provisional Death Counts for COVID-19" 

Male_DT <- My_data %>% 
  group_by(Age_Range, Year)  %>%
  filter(Sex == "Male") %>%
  summarise("Deaths with C19" = Deaths_with_Covid,
            "Deaths All Causes" = Death_All_Causes,
            "C19 to All Deaths (%)" = (C19_Ratio * 100) %>% round(2) ) 

create_dt(Male_DT)
```


### Key Stats

```{r}

cust_cap <-  "CDC Data: Provisional Death Counts for COVID-19"

RR_ALL_DT <- Comp_2021_vs_2020_ALL %>%
  mutate(across(where(is.numeric), round, 4), Per_Change = Per_Change * 100) 

create_dt(RR_ALL_DT)
```




```{r}
Comp_2021_vs_2020_ALL  %>%
  mutate(across(where(is.numeric), round, 4), Per_Change = Per_Change * 100) %>%
  reactable(., highlight = TRUE, striped = TRUE, 
            theme = reactableTheme(stripedColor = "#EDEDED",
                                   highlightColor = "#FFE4E1")) 
```



### Diff Plot

```{r}
per_change_both  <- Comp_2021_vs_2020 %>%
  ggplot(aes(x = Age_Range,  y = Per_Change,  fill =  Per_Change)) +
  geom_col() +
  facet_grid(Sex~ .) +
  scale_fill_viridis_c() +
  guides(fill = "none")  +
  labs(title = " 2021 vs 2020: % Change in Deaths with Covid",
       subtitle = "US CDC data for years 2021 & 2020",
       x = "Age Ranges", y = "Percentage Change",
      caption = "Data Humanist, CC BY-SA 4.0") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     breaks = breaks_per)

per_change_both  
```



```{r}
both_rr <- Comp_2021_vs_2020 %>%
  mutate(across(where(is.numeric), round, 2)) %>%
  ggplot(aes(x = Age_Range,  y = Relative_Risk,  fill =  Relative_Risk)) +
  geom_col() +
  facet_grid(Sex~ .) +
  scale_fill_viridis_c(option = "D") +
  guides(fill = "none")  +
  labs(title = "Relative Risk: Dying with Covid in 2021 vs. 2020",
       subtitle = "US CDC data for years 2021 & 2020",
       caption = "Data Humanist, CC BY-SA 4.0",
       x = "Age Ranges", y = "Relative Risk") +
  geom_hline(yintercept = 1, color = "white", lty = 3) +
  geom_text( aes(label = paste0(round({Relative_Risk}, 2), "x" ) ), size = 3, 
             vjust = 1.5, color= "red") 

both_rr
```


### Diff Stats

```{r}

cust_cap <-  "2021 vs 2020: Changes in Death Percentage and Relative Risk"

Comp_2021_vs_2020_DT <- Comp_2021_vs_2020 %>%
  mutate(across(where(is.numeric), round, 4), Per_Change = Per_Change * 100) 

create_dt(Comp_2021_vs_2020_DT)

```




### Raw Data Sets


```{r}
cust_cap <-  "Tidy Version: Long Format"


My_data_long %>% 
  mutate(across(where(is.numeric), round, 7)) %>% 
  create_dt()

All_Sexes_long%>% 
  mutate(across(where(is.numeric), round, 7)) %>% 
  create_dt()

```



